Does Anybody Really Know What Time It Is?\\ Automating the Extraction of Date Scalars

### 2015-07-09 nlp.tex
Text for grammar constraints

\begin{grammar}
<TimeGrammar> ::= <Hours> ':' <Minutes> ':' <Seconds> <AMPM>;\\

@@

<AMPM>  ::=  WORD('a.m.') 
			 \alt WORD('p.m.') ;

@@

@@

\subsection{Translation of Grammar Output to ICU Format}

@@

The EBNF date-time grammar includes a large number of syntactically correct but semantically invalid
date-time expressions. For example, there is no  `November 31, 2015', `February 29, 2013',
or `Sunday, May  5, 1965'. November only as 30 days in any year; 2013 was not a leap year; and May 5, 1965 was a Wednesday.

While custom production rules can be added to the existing grammar to exclude such expressions, this approach is not optimal as it leads to a rather large grammar that needs to account for every single semantically valid date-time sequence of terminal symbols. Rather, we define a grammar that excludes such expressions. By intersecting this constraint
grammar with the Date-Time grammar, we can define a language that contains only semantically valid dates.

>>>>

\begin{grammar}
<TimeGrammar> ::= <Hours> ':' <Minutes> ':' <Seconds> ;\\

@@

@@

<Day>     ::= NUMBER(1-31);

<Hour> ::= <TwelveHour> | <TwentyFourHour>

<TwelveHour> ::= NUMBER(1-12);

<TwentyFourHour> ::= NUMBER(1-24);

@@

\subsection{Translation to ICU Format}

@@

The EBNF date-time grammar includes a large number of syntactically correct but semantically invalid
date-time expressions. While we have added range restrictions to symbols such as \texttt{Hour} (1--12 for 12-hour format and 1--24 for 24-hour format), \texttt{Days} (1--7), \texttt{Month} (1--12), there are special cases that need to be accounted for. For example, there is no  `November 31, 2015', `February 29, 2013', or `Sunday, May  5, 1965'. November only as 30 days in any year; 2013 was not a leap year; and May 5, 1965 was a Wednesday.

While custom production rules can be added to the existing grammar to exclude such expressions, this approach is not optimal as it leads to a rather large grammar that needs to account for every single semantically valid date-time sequence of terminal symbols. Rather, we modify the existing grammar with the following additional constraints to the \texttt{Day} terminal symbol for excluding such expressions:
\begin{itemize}
\item Restriction on the distribution of $30$ and $31$:
Months usually alternate between lengths of 30 and 31 days. We can use x mod 2 to get an alternating pattern of 1 and 0, then just add our constant base number of days:

\begin{equation}
\texttt{Day} = 30 + x \Mod 2
\end{equation}

where $x \in [1..12]$ for each of the $12$ months.

Except February, Equation $1$ addresses January and March through July. After July, the pattern should skip one, and the rest of the months should follow the alternating pattern inversely.

To obtain an inverse pattern of alternating 0 and 1, we add 1 to the dividend:

\begin{equation}
\texttt{Day} = 30 + (x + 1) \Mod 2
\end{equation}

In Equation $2$ the number of days for August through December is correct ($x \in [8..12]$), but not for the remaining months. We hence introduce a bit-masking function so that the equation is equal to 1 over the desired domain and 0 otherwise. Multiplying a term by this expression will result in the term being cancelled out outside its domain. To mask the latter piece of our function, we need an expression equal to $1$ where $8 \le x\le 12$. Floor division by 8 works well, since $x < 16$.

Now if we substitute this expression in the x + 1 dividend of Equarion $2$, we can invert the pattern using our mask:

\begin{equation}
\texttt{Day} = 30 + (x + \floor{\frac{x}{8}}) \Mod 2
\end{equation}

Equation $3$ is now added as a restriction to the production rule \begin{grammar} <Days> ::= NUMBER(1-31) \end{grammar}.
\item LeapYear Restriction on February 29.

\end{itemize}
